---
title: "Schedule"
format:
  html:
    toc: true
    toc-depth: 4
sidebar: false
page-layout: full
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, include=FALSE}
library(tidyverse)
library(knitr)
```

## Course Schedule

Below is the schedule for the course. Note that this schedule may change as the course progresses.


::: {.callout-note}
## Schedule Notes

1. All deadlines are at 11:59 PM Pacific Time on the specified date
2. Schedule is subject to change based on course needs
3. Additional readings and resources will be posted on the course website
:::

```{r packages-data, include=FALSE }
library(tidyverse)
library(knitr)
library(kableExtra)

withr::with_dir(here::here(), {
  schedule_nested <- targets::tar_read(schedule_page_data)
  targets::tar_load(schedule_ical_file)
})

base_url <- yaml::read_yaml(here::here("_quarto.yml"))$website$`site-url`
ical_url_absolute <- paste0(base_url, "/", schedule_ical_file)
```

::: {.schedule-page}

Here's your roadmap for the semester!

- [**Content**](/content/) ({{< fa book-open-reader >}}): This page contains the recorded lectures and slides for the week. Watch the lectures **before** our in-person class.

- [**Assignment**](/assignment/) ({{< fa pen-ruler >}}): This page contains the instructions for each assignment. 


```{r build-table, include=FALSE}
show_table <- function(group_id) {
  # Add a heading
  cat(as.character(paste("\n\n###", schedule_nested$group[[group_id]], "\n\n")))
  
  # Make the table
  tbl <- schedule_nested$data[[group_id]] %>% 
    select(-subgroup) %>% 
    kbl(escape = FALSE, align = "rlccc", table.attr = 'class="schedule-table"') %>% 
    kable_styling() %>% 
    column_spec(1, width = "20%", extra_css = "padding-right: 20px;") %>% 
    column_spec(2, width = "50%") %>% 
    column_spec(3:4, width = "10%") %>% 
    pack_rows(index = schedule_nested$subgroup_index[[group_id]],
              label_row_css = "border-bottom: 2px solid #000000;")
  
  cat(tbl)
}
```

```{r show-table, echo=FALSE, results="asis"}
walk(seq(1, nrow(schedule_nested)), ~show_table(.x))
```


