---
title: "Schedule"
format:
  html:
    toc: true
    toc-depth: 4
sidebar: false
page-layout: full
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, include=FALSE}
library(tidyverse)
library(knitr)
```

## Course Schedule

Below is the schedule for the course. Note that this schedule may change as the course progresses.

### Week 1: Course Introduction

- **Session 1**: Course Overview
  - Topics: Course objectives, requirements, and expectations; Bayes' Rule

- **Session 2**: Gardens of forking data
  - Topics: Bayes as counting
  - [Assignment 1](/assignment/01-problem-set.html) released

### Week 2: Topic 1

- **Session 3**: Topic 1.1
  - Topics: Key concepts and methods
  - [Slides](/slides/02-slides.html) ([PDF](/slides/02-slides.pdf))

- **Session 4**: Topic 1.2
  - Topics: Applications and examples
  - [Assignment 1](/assignment/01-problem-set.html) due

### Week 3: Topic 2

- **Session 5**: Topic 2.1
  - Topics: Key concepts and methods
  - [Slides](/slides/03-slides.html) ([PDF](/slides/03-slides.pdf))

- **Session 6**: Topic 2.2
  - Topics: Applications and examples
  - [Assignment 2](/assignment/02-problem-set.html) released

::: {.callout-note}
## Schedule Notes

1. All deadlines are at 11:59 PM Pacific Time on the specified date
2. Schedule is subject to change based on course needs
3. Additional readings and resources will be posted on the course website
:::

```{r packages-data, include=FALSE, eval = FALSE }
library(tidyverse)
library(knitr)
library(kableExtra)

# withr::with_dir(here::here(), {
#   schedule_nested <- targets::tar_read(schedule_page_data)
#   targets::tar_load(schedule_ical_file)
# })

base_url <- yaml::read_yaml(here::here("_quarto.yml"))$website$`site-url`
# ical_url_absolute <- paste0(base_url, "/", schedule_ical_file)
```

::: {.schedule-page}

Here's your roadmap for the semester!

- [**Content**](/content/) ({{< fa book-open-reader >}}): This page contains the recorded lectures and slides for the week. Watch the lectures **before** our in-person class.

- [**Assignment**](/assignment/) ({{< fa pen-ruler >}}): This page contains the instructions for each assignment. 
