---
title: "Introduction to the Course"
author: "{{< var instructor.name >}}"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```

```{r packages-data, include=FALSE}
library(tidyverse)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-7

# Introduction to<br>the Course

.class-info[

**Session 1**

.light[{{< var course.number >}}: {{< var course.title >}}<br>
{{< var university.department >}}
]

]

---

name: outline
class: title title-inv-8

# Plan for today

--

.box-5.medium.sp-after-half[Course overview]

--

.box-3.medium.sp-after-half[Course requirements]

--

.box-6.medium.sp-after-half[Getting started]

---

layout: false
name: overview
class: center middle section-title section-title-5 animated fadeIn

# Course overview

---

class: title title-5

# Course objectives

.box-inv-5.medium[After completing this course, you will be able to:]

--

.box-5[Objective 1]

--

.box-5[Objective 2]

--

.box-5[Objective 3]

---

# Course structure

.box-inv-5.medium[Weekly topics]

.box-5[Topic descriptions]

.box-inv-5.medium[Assignments]

.box-5[Assignment descriptions]

---

layout: false
name: requirements
class: center middle section-title section-title-3 animated fadeIn

# Course requirements

---

class: title title-3

# Required materials

.pull-left[
.box-inv-3[Textbook 1]

.box-3[Description]
]

.pull-right[
.box-inv-3[Textbook 2]

.box-3[Description]
]

---

# Software requirements

.center[
.box-inv-3.medium[Required software and tools]
]

.pull-left[
.box-3[Software 1]

.box-3[Software 2]
]

.pull-right[
.box-3[Software 3]

.box-3[Software 4]
]

---

layout: false
name: getting-started
class: center middle section-title section-title-6 animated fadeIn

# Getting started

---

class: title title-6

# Next steps

.box-inv-6.medium[What to do before next class]

--

.box-6[Step 1]

--

.box-6[Step 2]

--

.box-6[Step 3]

---

# Questions?

.center[
.box-6.huge[Thank you!]
]
