---
title: "week 7: multilevel models"
subtitle: "multilevel adventures"
format: 
  revealjs:
    css: xaringan-themer2.css
    nature:
      highlightStyle: solarized-dark
      highlightLines: true
      countIncrementalSlides: false
      mathjax: "default"
    self-contained: false  # Ensures correct embedding
    embed-resources: true  # Embeds required assets
    slide-number: true
    code-annotations: hover
execute:
  echo: false        
---

```{r, message = F, warning = F}
library(tidyverse)
library(psych)
library(cowplot)
library(patchwork)
library(here)
library(brms) 
library(tidybayes) 
```

```{r, echo = F}
knitr::opts_chunk$set(fig.retina=3, echo=TRUE)
theme_set(theme_cowplot())
default_palettes <- list(
  c("#5e8485" , "#0f393a") ,
  c("#1c5253" , "#5e8485" , "#0f393a") , 
  # palette with 5 colours
 c( "#1c5253" , "#e07a5f", "#f2cc8f" , "#81b29a" , "#3d405b" ) ,
  # same palette interpolated to 8 colours
 c( "#1c5253" , "#e07a5f", "#f2cc8f" , "#81b29a" , "#3d405b" , "#a7a844" , "#69306d" ) 
  
)

options(ggplot2.discrete.fill = default_palettes, 
        ggplot2.discrete.colour = default_palettes)
```

## slopes

Let's start by simulating the cafe data.

```{r}
# ---- set population-level parameters -----
a <- 3.5       # average morning wait time
b <- (-1)      # average difference afternoon wait time
sigma_a <- 1   # std dev in intercepts
sigma_b <- 0.5 # std dev in slopes
rho <- (-0.7)  #correlation between intercepts and slopes

# ---- create vector of means ----
Mu <- c(a, b)

# ---- create matrix of variances and covariances ----
sigmas <- c(sigma_a,sigma_b) # standard deviations
Rho <- matrix( c(1,rho,rho,1) , nrow=2 ) # correlation matrix
# now matrix multiply to get covariance matrix
Sigma <- diag(sigmas) %*% Rho %*% diag(sigmas)

# ---- simulate intercepts and slopes -----
N_cafes = 20
library(MASS)
set.seed(5)
vary_effects <- mvrnorm( n=N_cafes, mu = Mu, Sigma=Sigma)
a_cafe <- vary_effects[, 1]
b_cafe <- vary_effects[, 2]

# ---- simulate observations -----

set.seed(22)
N_visits <- 10
afternoon <- rep(0:1,N_visits*N_cafes/2)
cafe_id <- rep( 1:N_cafes , each=N_visits )
mu <- a_cafe[cafe_id] + b_cafe[cafe_id]*afternoon
sigma <- 0.5 # std dev within cafes
wait <- rnorm( N_visits*N_cafes , mu , sigma )
d <- data.frame( cafe=cafe_id , afternoon=afternoon , wait=wait )

```

------------------------------------------------------------------------

### a simulation note from RM

> In this exercise, we are simulating data from a generative process and then analyzing that data with a model that reflects exactly the correct structure of that process. But in the real world, we’re never so lucky. Instead we are always forced to analyze data with a model that is MISSPECIFIED: The true data-generating process is different than the model. Simulation can be used however to explore misspecification. Just simulate data from a process and then see how a number of models, none of which match exactly the data-generating process, perform. And always remember that Bayesian inference does not depend upon data-generating assumptions, such as the likelihood, being true. Non-Bayesian approaches may depend upon sampling distributions for their inferences, but this is not the case for a Bayesian model. In a Bayesian model, a likelihood is a prior for the data, and inference about parameters can be surprisingly insensitive to its details.

------------------------------------------------------------------------

**Mathematical model:**

likelihood function and linear model

\begin{align*}
W_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{CAFE[i]} + \beta_{CAFE[i]}A_i
\end{align*}

varying intercepts and slopes

\begin{align*}
\begin{bmatrix} \alpha_{CAFE[i]} \\ \beta_{CAFE[i]} \end{bmatrix} &\sim \text{MVNormal}( \begin{bmatrix} \alpha \\ \beta \end{bmatrix}, \mathbf{S}) \\
\mathbf{S} &\sim \begin{pmatrix} \sigma_{\alpha}, & 0 \\ 0, & \sigma_{\beta}\end{pmatrix}\mathbf{R}\begin{pmatrix} \sigma_{\alpha}, & 0 \\ 0, & \sigma_{\beta}\end{pmatrix} \\
\end{align*}

priors

\begin{align*}
\alpha &\sim \text{Normal}(5,2) \\
\beta &\sim  \text{Normal}(-1,0.5) \\
\sigma &\sim  \text{Exponential}(1) \\
\sigma_{\alpha} &\sim  \text{Exponential}(1) \\
\sigma_{\beta} &\sim  \text{Exponential}(1) \\
\mathbf{R} &\sim \text{LKJcorr}(2)
\end{align*}

------------------------------------------------------------------------

### LKJ correlation prior

```{r}
#| code-fold: true
# examples
rlkj_1 = rethinking::rlkjcorr(1e4, K=2, eta=1)
rlkj_2 = rethinking::rlkjcorr(1e4, K=2, eta=2)
rlkj_4 = rethinking::rlkjcorr(1e4, K=2, eta=4)
rlkj_6 = rethinking::rlkjcorr(1e4, K=2, eta=6)
data.frame(rlkj_1= rlkj_1[,1,2], 
           rlkj_2= rlkj_2[,1,2], 
           rlkj_4= rlkj_4[,1,2],
           rlkj_6= rlkj_6[,1,2]) %>% 
  ggplot() +
  geom_density(aes(x=rlkj_1, color = "1"), alpha=.3) +
  geom_density(aes(x=rlkj_2, color = "2"), alpha=.3) +
  geom_density(aes(x=rlkj_4, color = "4"), alpha=.3) +
  geom_density(aes(x=rlkj_6, color = "6"), alpha=.3) +
  labs(x="R", color="eta") +
  theme(legend.position = "top")
```

------------------------------------------------------------------------

```{r}
m1 <- brm(
  data = d,
  family = gaussian,
  wait ~ 1 + afternoon + (1 + afternoon | cafe),
  prior = c(
    prior( normal(5,2),    class=Intercept ), 
    prior( normal(-1, .5), class=b),
    prior( exponential(1), class=sd),
    prior( exponential(1), class=sigma),
    prior( lkj(2),         class=cor)
  ), 
  iter=2000, warmup=1000, chains=4, cores=4, seed=9,
  file=here("files/models/72.1")
)
```

------------------------------------------------------------------------

```{r}
posterior_summary(m1) %>% round(3)
```

------------------------------------------------------------------------

Let's get the slopes and intercepts for each cafe.

```{r}
#| code-fold: true
intercepts = coef(m1)$cafe[ ,, "Intercept"]
slopes = coef(m1)$cafe[,, "afternoon"]
cafe_params = data.frame(
  cafe=1:20,
  intercepts=intercepts[, 1],
  slopes=slopes[, 1]
) 
cafe_params %>% round(3)
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  ggplot( aes(x=intercepts, y=slopes) ) +
  geom_point(size = 2) 

```

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  ggplot( aes(x=intercepts, y=slopes) ) +
  stat_ellipse() +
  geom_point(size = 2) 

```

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  ggplot( aes(x=intercepts, y=slopes) ) +
  mapply(function(level) {
    stat_ellipse(geom  = "polygon", type = "norm",
                 linewidth = 0, alpha = .1, fill = "#1c5253",
                 level = level)
    }, 
    # enter the levels here
    level = c(1:9 / 10, .99)) +
  geom_point(size = 2) 

```

More about `stat_ellipse` [here](https://ggplot2.tidyverse.org/reference/stat_ellipse.html).

------------------------------------------------------------------------

### exercise

Now use the slopes and intercepts to calculate the expected morning and afternoon wait times for each cafe. Plot these as a scatterplot. Bonus for ellipses.

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  mutate(
    morning = intercepts, 
    afternoon = intercepts + slopes
  ) %>% 
  ggplot( aes(x=morning, y=afternoon) ) +
  mapply(function(level) {
    stat_ellipse(geom  = "polygon", type = "norm",
                 linewidth = 0, alpha = .1, fill = "#1c5253",
                 level = level)
    }, 
    # enter the levels here
    level = c(1:9 / 10, .99)) +
  geom_point(size = 2)+
  labs(x="morning wait time",
       y="afternoon wait time")
```

------------------------------------------------------------------------

What is the covariance of our intercepts and slopes?

```{r}
#| code-fold: true
post = as_draws_df(m1)
rlkj_2 = rethinking::rlkjcorr(nrow(post), K=2, eta=2)

data.frame(prior= rlkj_2[,1,2],
           posterior = post$cor_cafe__Intercept__afternoon) %>% 
  ggplot() +
  geom_density(aes(x=prior, color = "prior"), alpha=.3) +
  geom_density(aes(x=posterior, color = "posterior"), alpha=.3) +
  scale_color_manual(values = c("#1c5253" , "#e07a5f")) +
  labs(x="R") +
  theme(legend.position = "top")
```

------------------------------------------------------------------------

## predictors at different levels

Returning to our personality data from Tuesday.

```{r}
data_path = "https://raw.githubusercontent.com/sjweston/uobayes/refs/heads/main/files/data/external_data/mlm.csv"
d <- read.csv(data_path)

d %>% count(wave)
d %>% count(group)
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
d %>% 
  ggplot(aes(x=wave, y=con, group=id)) +
  geom_line(alpha=.3) +
  facet_wrap(~group)
```

`wave` is a within-person variable (Level 1), whereas `group` is a between-person variable (Level 2).

------------------------------------------------------------------------

**Mathematical model with level-2 covariate**

::::: columns
::: {.column width="50%"}

Likelihood function and linear model

\begin{align*}
\text{con}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{\text{id}[i]} + \beta_{\text{id}[i]}t_i
\end{align*}

Varying intercepts and slopes with level-2 predictor:

\begin{align*}
\alpha_{\text{id}[i]} &= \alpha + \gamma_\alpha G_{\text{id}[i]} + u_{\alpha,\text{id}[i]} \\
\beta_{\text{id}[i]} &= \beta + \gamma_\beta G_{\text{id}[i]} + u_{\beta,\text{id}[i]}
\end{align*}

Random effects:

\begin{align*}
\begin{bmatrix} u_{\alpha,\text{id}[i]} \\ u_{\beta,\text{id}[i]} \end{bmatrix} &\sim \text{MVNormal}\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \mathbf{S}\right) \\
\mathbf{S} &= \begin{pmatrix} \sigma_{\alpha} & 0 \\ 0 & \sigma_{\beta}\end{pmatrix}\mathbf{R}\begin{pmatrix} \sigma_{\alpha} & 0 \\ 0 & \sigma_{\beta}\end{pmatrix}
\end{align*} 

:::

::: {.column width="50%"}


Priors: 
\begin{align*}
\alpha &\sim \text{Normal}(0,1) \\
\beta &\sim \text{Normal}(0,1) \\
\gamma_\alpha &\sim \text{Normal}(0,1) \\
\gamma_\beta &\sim \text{Normal}(0,1) \\
\sigma &\sim \text{Exponential}(1) \\
\sigma_{\alpha} &\sim \text{Exponential}(1) \\
\sigma_{\beta} &\sim \text{Exponential}(1) \\
\mathbf{R} &\sim \text{LKJcorr}(2)
\end{align*}

:::
:::::


------------------------------------------------------------------------

```{r}
m2 <- brm(
  data=d,
  family=gaussian,
  con ~ 1 + time + group + time:group + (1 + time | id),
  prior = c( prior( normal(0,1),    class=Intercept ),
             prior( normal(0,1),    class=b ),
             prior( exponential(1), class=sigma ),
             prior( exponential(1), class=sd ),
             prior( lkj(2),         class=cor)),
  iter=4000, warmup=1000, seed=9, cores=4, chains=4,
  file=here("files/models/72.2")
)
```

---

```{r}
m2
```

---

```{r}
m2 <- brm(
  data=d,
  family=gaussian,
  bf(con ~ 0 + a + b*time + aid + bid*time,
     a ~ 0 + group,
     b ~ 0 + group,
     aid ~ 1 + (1 |id),
     bid ~ 1 + (1 |id),
     nl=TRUE),
  prior = c( prior( normal(0,1),    class=b, nlpar=a ),
             prior( normal(0,1),    class=b, nlpar=b ),
             prior( exponential(1), class=sigma )),
  iter=4000, warmup=1000, seed=9, cores=4, chains=4,
  file=here("files/models/72.2b")
)
```

---

```{r}
m2
```
