---
title: "week 8: multilevel models"
subtitle: "adventures in covariance"
format: 
  revealjs:
    css: xaringan-themer2.css
    nature:
      highlightStyle: solarized-dark
      highlightLines: true
      countIncrementalSlides: false
      mathjax: "default"
    self-contained: false  # Ensures correct embedding
    embed-resources: true  # Embeds required assets
    slide-number: true
    code-annotations: hover
execute:
  echo: true        
---

```{r, message = F, warning = F, echo= T}
library(tidyverse)
library(psych)
library(cowplot)
library(patchwork)
library(here)
library(brms) 
library(tidybayes) 
```

[Download (almost) all model files here.](https://github.com/sjweston/uobayes/raw/refs/heads/main/files/share/mods7-2.zip)


```{r, echo = F}
knitr::opts_chunk$set(fig.retina=3, echo=TRUE)
theme_set(theme_cowplot())
default_palettes <- list(
  c("#5e8485" , "#0f393a") ,
  c("#1c5253" , "#5e8485" , "#0f393a") , 
  # palette with 5 colours
 c( "#1c5253" , "#e07a5f", "#f2cc8f" , "#81b29a" , "#3d405b" ) ,
  # same palette interpolated to 8 colours
 c( "#1c5253" , "#e07a5f", "#f2cc8f" , "#81b29a" , "#3d405b" , "#a7a844" , "#69306d" ) 
  
)

options(ggplot2.discrete.fill = default_palettes, 
        ggplot2.discrete.colour = default_palettes)
```


## today's topics

  * Varying slopes

---


## varying slopes

Let's start by simulating cafe data.

```{r}
# ---- set population-level parameters -----
a <- 3.5       # average morning wait time
b <- (-1)      # average difference afternoon wait time
sigma_a <- 1   # std dev in intercepts
sigma_b <- 0.5 # std dev in slopes
rho <- (-0.7)  #correlation between intercepts and slopes

# ---- create vector of means ----
Mu <- c(a, b)

# ---- create matrix of variances and covariances ----
sigmas <- c(sigma_a,sigma_b) # standard deviations
Rho <- matrix( c(1,rho,rho,1) , nrow=2 ) # correlation matrix
# now matrix multiply to get covariance matrix
Sigma <- diag(sigmas) %*% Rho %*% diag(sigmas)

# ---- simulate intercepts and slopes -----
N_cafes = 20
library(MASS)
set.seed(5)
vary_effects <- mvrnorm( n=N_cafes, mu = Mu, Sigma=Sigma)
a_cafe <- vary_effects[, 1]
b_cafe <- vary_effects[, 2]

# ---- simulate observations -----

set.seed(22)
N_visits <- 10
afternoon <- rep(0:1,N_visits*N_cafes/2)
cafe_id <- rep( 1:N_cafes , each=N_visits )
mu <- a_cafe[cafe_id] + b_cafe[cafe_id]*afternoon
sigma <- 0.5 # std dev within cafes
wait <- rnorm( N_visits*N_cafes , mu , sigma )
d3 <- data.frame( cafe=cafe_id , afternoon=afternoon , wait=wait )

```

---

```{r}
rethinking::precis(d3)
d3 %>% filter(cafe==1)
```

------------------------------------------------------------------------

### a simulation note from RM

> In this exercise, we are simulating data from a generative process and then analyzing that data with a model that reflects exactly the correct structure of that process. But in the real world, we’re never so lucky. Instead we are always forced to analyze data with a model that is MISSPECIFIED: The true data-generating process is different than the model. Simulation can be used however to explore misspecification. Just simulate data from a process and then see how a number of models, none of which match exactly the data-generating process, perform. And always remember that Bayesian inference does not depend upon data-generating assumptions, such as the likelihood, being true. Non-Bayesian approaches may depend upon sampling distributions for their inferences, but this is not the case for a Bayesian model. In a Bayesian model, a likelihood is a prior for the data, and inference about parameters can be surprisingly insensitive to its details.

------------------------------------------------------------------------

**Mathematical model:**

likelihood function and linear model

\begin{align*}
W_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{CAFE[i]} + \beta_{CAFE[i]}(\text{Afternoon}_i)
\end{align*}

varying intercepts and slopes

\begin{align*}
\begin{bmatrix} \alpha_{CAFE[i]} \\ \beta_{CAFE[i]} \end{bmatrix} &\sim \text{MVNormal}( \begin{bmatrix} \alpha \\ \beta \end{bmatrix}, \mathbf{S}) \\
\mathbf{S} &\sim \begin{pmatrix} \sigma_{\alpha}, & 0 \\ 0, & \sigma_{\beta}\end{pmatrix}\mathbf{R}\begin{pmatrix} \sigma_{\alpha}, & 0 \\ 0, & \sigma_{\beta}\end{pmatrix} \\
\end{align*}

priors

\begin{align*}
\alpha &\sim \text{Normal}(5,2) \\
\beta &\sim  \text{Normal}(-1,0.5) \\
\sigma &\sim  \text{Exponential}(1) \\
\sigma_{\alpha} &\sim  \text{Exponential}(1) \\
\sigma_{\beta} &\sim  \text{Exponential}(1) \\
\mathbf{R} &\sim \text{LKJcorr}(2)
\end{align*}

------------------------------------------------------------------------

### LKJ correlation prior

```{r}
#| code-fold: true
# examples
rlkj_1 = rethinking::rlkjcorr(1e4, K=2, eta=1)
rlkj_2 = rethinking::rlkjcorr(1e4, K=2, eta=2)
rlkj_4 = rethinking::rlkjcorr(1e4, K=2, eta=4)
rlkj_6 = rethinking::rlkjcorr(1e4, K=2, eta=6)
data.frame(rlkj_1= rlkj_1[,1,2], 
           rlkj_2= rlkj_2[,1,2], 
           rlkj_4= rlkj_4[,1,2],
           rlkj_6= rlkj_6[,1,2]) %>% 
  ggplot() +
  geom_density(aes(x=rlkj_1, color = "1"), alpha=.3) +
  geom_density(aes(x=rlkj_2, color = "2"), alpha=.3) +
  geom_density(aes(x=rlkj_4, color = "4"), alpha=.3) +
  geom_density(aes(x=rlkj_6, color = "6"), alpha=.3) +
  labs(x="R", color="eta") +
  theme(legend.position = "top")
```

------------------------------------------------------------------------

```{r}
m5 <- brm(
  data = d3,
  family = gaussian,
  wait ~ 1 + afternoon + (1 + afternoon | cafe),
  prior = c(
    prior( normal(5,2),    class=Intercept ), 
    prior( normal(-1, .5), class=b),
    prior( exponential(1), class=sd),
    prior( exponential(1), class=sigma),
    prior( lkj(2),         class=cor)
  ), 
  iter=2000, warmup=1000, chains=4, cores=4, seed=9,
  file=here("files/models/m72.5")
)
```

------------------------------------------------------------------------

```{r}
posterior_summary(m5) %>% round(2)
```

------------------------------------------------------------------------

Let's get the slopes and intercepts for each cafe.

```{r}
#| code-fold: true

cafe_params = m5 %>% spread_draws(b_Intercept, b_afternoon, r_cafe[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_cafe) %>% 
  mutate( intercepts =b_Intercept + Intercept,
          slopes = b_afternoon + afternoon) %>% 
  mean_qi(intercepts, slopes) 

cafe_params %>% 
  dplyr::select(cafe=id, intercepts, slopes) %>% 
  round(2)
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  ggplot( aes(x=intercepts, y=slopes) ) +
  geom_point(size = 2) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  ggplot( aes(x=intercepts, y=slopes) ) +
  stat_ellipse() +
  geom_point(size = 2) 

```

------------------------------------------------------------------------

```{r}
#| code-fold: true
cafe_params %>% 
  ggplot( aes(x=intercepts, y=slopes) ) +
  mapply(function(level) {
    stat_ellipse(geom  = "polygon", type = "norm",
                 linewidth = 0, alpha = .1, fill = "#1c5253",
                 level = level)
    }, 
    # enter the levels here
    level = c(1:9 / 10, .99)) +
  geom_point(size = 2) 

```

More about `stat_ellipse` [here](https://ggplot2.tidyverse.org/reference/stat_ellipse.html).

------------------------------------------------------------------------

We can use the slopes and intercepts to calculate the expected morning and afternoon wait times for each cafe. 


```{r}
#| code-fold: true
cafe_params %>% 
  mutate(
    morning = intercepts, 
    afternoon = intercepts + slopes
  ) %>% 
  ggplot( aes(x=morning, y=afternoon) ) +
  mapply(function(level) {
    stat_ellipse(geom  = "polygon", type = "norm",
                 linewidth = 0, alpha = .1, fill = "#1c5253",
                 level = level)
    }, 
    # enter the levels here
    level = c(1:9 / 10, .99)) +
  geom_point(size = 2)+
  labs(x="morning wait time",
       y="afternoon wait time")
```

---

Or, we can use `epred_draws`.

```{r}
#| code-fold: true
d3 %>% 
  distinct(cafe, afternoon) %>% 
  add_epred_draws(m5) %>% 
  group_by(cafe, afternoon) %>% 
  summarise(wait = mean(.epred), .groups = "drop") %>% 
  mutate(afternoon =ifelse(afternoon==1, "afternoon", "morning")) %>% 
  pivot_wider(names_from = afternoon, values_from=wait) %>% 
  ggplot( aes(x=morning, y=afternoon) ) +
  mapply(function(level) {
    stat_ellipse(geom  = "polygon", type = "norm",
                 linewidth = 0, alpha = .1, fill = "#1c5253",
                 level = level)
    }, 
    # enter the levels here
    level = c(1:9 / 10, .99)) +
  geom_point(size = 2)+
  labs(x="morning wait time",
       y="afternoon wait time")

```


------------------------------------------------------------------------

What is the correlation of our intercepts and slopes?

```{r}
m5 %>% spread_draws(cor_cafe__Intercept__afternoon) %>% mean_qi()
```


```{r}
#| code-fold: true
post = as_draws_df(m5)
rlkj_2 = rethinking::rlkjcorr(nrow(post), K=2, eta=2)

data.frame(prior= rlkj_2[,1,2],
           posterior = post$cor_cafe__Intercept__afternoon) %>% 
  ggplot() +
  geom_density(aes(x=prior, color = "prior"), alpha=.3) +
  geom_density(aes(x=posterior, color = "posterior"), alpha=.3) +
  scale_color_manual(values = c("#1c5253" , "#e07a5f")) +
  labs(x="R") +
  theme(legend.position = "top")
```

------------------------------------------------------------------------

## multilevel people again

Let's apply what we've learned to our affect data: 


```{r}
data_path = "https://raw.githubusercontent.com/sjweston/uobayes/refs/heads/main/files/data/external_data/williams.csv"
d <- read.csv(data_path)
rethinking::precis(d)
```

---

Here's our original intercept-only model (for comparison).

\begin{align*}
\text{PA.std}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{\text{id}[i]} \\
\alpha_j &\sim \text{Normal}(\bar{\alpha}, \sigma_{\alpha}) \text{ for j in 1...239}\\
\bar{\alpha} &\sim \text{Normal}(0, 1.5)\\
\sigma_{\alpha} &\sim \text{Exponential}(1) \\
\sigma &\sim \text{Exponential}(1) \\
\end{align*}

```{r}
m1 <- brm(
  data=d,
  family=gaussian,
  PA.std ~ 1 + (1 | id), 
  prior = c( prior(normal(0, 1.5), class=Intercept),
             prior(exponential(1), class=sd),
             prior(exponential(1), class=sigma)),
  iter=4000, warmup=1000, chains=4, cores=4, seed=9, #running this longer so it mixes
  control = list(adapt_delta =.99),
  file=here("files/models/m71.2")
)
```

---

```{r}
#| code-fold: true

seed=4
set.seed(seed)
p1 = d %>% 
  nest(data = -id) %>% 
  sample_n(size = 3) %>% 
  unnest() %>% 
  ggplot(aes(x=day, y=PA.std, group=id)) +
  geom_line(alpha=.3) +
  facet_wrap(~id, nrow=1) +
  labs(x="day", y="PA")

set.seed(seed) 
p2 = d %>% 
  nest(data = -id) %>% 
  sample_n(size = 3) %>% 
  unnest() %>% 
  ggplot(aes(x=PA_lag, y=PA.std, group=id)) +
  geom_line(alpha=.3) +
  facet_wrap(~id, nrow=1) +
  labs(x="PA yesterday", y="PA today")

p1/p2
```

------------------------------------------------------------------------

**Mathematical model with varying slopes**

::::: columns
::: {.column width="50%"}

Likelihood function and linear model

\begin{align*}
\text{PA}_{ij} &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{\text{id}[i]} + \beta_{\text{id}[i]}\text{PA}_{i,t-1}
\end{align*}

Varying intercepts and slopes:

\begin{align*}
\alpha_{\text{id}[i]} &= \gamma_{\alpha} + U_{\alpha,\text{id}[i]} \\
\beta_{\text{id}[i]} &= \gamma_{\beta} + U_{\beta,\text{id}[i]}
\end{align*}

Random effects:

\begin{align*}
\begin{bmatrix} U_{\alpha,\text{id}[i]} \\ U_{\beta,\text{id}[i]} \end{bmatrix} &\sim \text{MVNormal}\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \mathbf{S}\right) \\
\mathbf{S} &= \begin{pmatrix} \sigma_{\alpha} & 0 \\ 0 & \sigma_{\beta}\end{pmatrix}\mathbf{R}\begin{pmatrix} \sigma_{\alpha} & 0 \\ 0 & \sigma_{\beta}\end{pmatrix}
\end{align*} 

:::

::: {.column width="50%"}


Priors: 
\begin{align*}
\gamma_\alpha &\sim \text{Normal}(0,1) \\
\gamma_\beta &\sim \text{Normal}(0,1) \\
\sigma &\sim \text{Exponential}(1) \\
\sigma_{\alpha} &\sim \text{Exponential}(1) \\
\sigma_{\beta} &\sim \text{Exponential}(1) \\
\mathbf{R} &\sim \text{LKJcorr}(2)
\end{align*}

:::
:::::


------------------------------------------------------------------------

```{r}
m6 <- brm(
  data=d,
  family=gaussian,
  PA.std ~ 1 + PA_lag + (1 + PA_lag | id),
  prior = c( prior( normal(0,1),    class=Intercept ),
             prior( normal(0,1),    class=b ),
             prior( exponential(1), class=sigma ),
             prior( exponential(1), class=sd ),
             prior( lkj(2),         class=cor)),
  iter=4000, warmup=1000, seed=9, cores=4, chains=4,
  file=here("files/models/m72.6")
)
```

---

```{r}
m6
```

---

What can we learn from this model?

  * Overall, what's the relationship between yesterday's PA and today's PA?
  * How much within-person variability is accounted for by holdover PA?
  * To what extent do people differ in the association of holdover PA?
  * Is someone's overall positive affect related to their holdover effect?
  
---

### Overall relationship

```{r}
set.seed(3)
post = as_draws_df(m6) 
# estimates from the posterior
post %>% mean_qi(b_PA_lag)
```

---

```{r}
#| code-fold: true
# plot posterior density
p1 = post %>% 
  ggplot(aes(x = b_PA_lag)) +
  stat_halfeye(fill="#1c5253") +
  geom_vline( aes(xintercept=0), linetype = "dashed" ) +
  labs( title="posterior density",
        x="slope",
        y=NULL) +
  scale_y_continuous(breaks=NULL)
# posterior lines
p2 = d %>% 
  sample_n(2000) %>% 
  ggplot(aes( x=PA_lag, y=PA.std )) +
  geom_point(alpha=.2, shape = 20) +
  geom_abline( aes(intercept=b_Intercept, slope=b_PA_lag),
               data=post[1:50, ],
               alpha=.3,
               color="#1c5253") +
  labs( x="NA",
        y="PA",
        title="posterior lines")

p1 + p2  
```


---

### Accounting for variability

How much within-person variability is accounted for by variability in negative affect? Let's compare estimates of the residual variability of these models.

```{r}
# model with no predictors
m1_sigma = spread_draws(m1, sigma) %>% mutate(model = "intercept only")
m6_sigma = spread_draws(m6, sigma) %>% mutate(model = "with lag")

m1_sigma %>% 
  full_join(m6_sigma) %>% 
  group_by(model) %>% 
  mean_qi(sigma)
```

---

```{r}
#| code-fold: true
p1 = m1_sigma %>% 
  full_join(m6_sigma) %>% 
  ggplot( aes(y=sigma, fill=model )) +
  stat_halfeye(alpha=.5) + 
  labs(
    x=NULL,
    y = "within-person sd",
    title = "posterior distributions of\neach model") +
  scale_x_continuous(breaks=NULL) +
  theme(legend.position = "bottom")

p2 = data.frame(diff = m6_sigma$sigma - m1_sigma$sigma) %>% 
  ggplot( aes(x=diff)) +
  stat_halfeye() + 
  labs(
    y=NULL,
    x = "m6-m1",
    title = "posterior distribution of\ndifference in sigma") +
  scale_y_continuous(breaks=NULL) 

p2 + p1
```

---

### individual differences

To what extent do people differ in the association of holdover PA?

```{r}
#| code-fold: true
seed = 2
set.seed(seed)
sample_id = sample(unique(d$id), replace=F, size = 6)
d %>% 
  filter(id %in% sample_id) %>% 
  ggplot( aes( x=day, y=PA.std ) ) +
  geom_point() +
  geom_line() +
  facet_wrap(~id) +
  labs(y="positive affect")
```

---

To what extent do people differ in the association of holdover PA?

```{r}
#| code-fold: true
d %>% 
  filter(id %in% sample_id) %>% 
  ggplot( aes( x=PA_lag, y=PA.std ) ) +
  geom_point() +
  facet_wrap(~id) +
  labs(y="today",
       x="yesterday")
```


---

To what extent do people differ in the association of holdover PA?

```{r}
#| code-fold: true
d %>% 
  filter(id %in% sample_id) %>% 
  ggplot( aes( x=PA_lag, y=PA.std ) ) +
  geom_point(alpha=.5, shape=20) +
  geom_abline( aes(intercept=b_Intercept, slope=b_PA_lag),
               data = post[1:50, ],
               alpha=.4, 
               color="#1c5253") +
  facet_wrap(~id) +
  labs(y="today",
       x="yesterday")
```

---

```{r}
m6 %>% spread_draws(b_Intercept, b_PA_lag, r_id[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_id) %>% 
  filter(id %in% sample_id)
```

---

```{r}
#| code-fold: true
#| 
post2 = m6 %>% spread_draws(b_Intercept, b_PA_lag, r_id[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_id) %>% 
  mutate(intercept = b_Intercept + Intercept,
         slope = b_PA_lag + PA_lag) %>% 
  filter(id %in% sample_id) %>% 
  with_groups(id, sample_n, 50)

d %>% 
  filter(id %in% sample_id) %>% 
  ggplot( aes( x=PA_lag, y=PA.std ) ) +
  geom_point(alpha=.5, shape=20) +
  geom_abline( aes(intercept=intercept, slope=slope, color=as.factor(id)),
               data = post2,
               alpha=.4) +
  facet_wrap(~id) +
  guides(color="none") +
  labs(y="today",
       x="yesterday")
```


---

```{r}
#| code-fold: true
#| 
post2 = m6 %>% spread_draws(b_Intercept, b_PA_lag, r_id[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_id) %>% 
  mutate(intercept = b_Intercept + Intercept,
         slope = b_PA_lag + PA_lag) %>% 
  filter(id %in% sample_id) %>% 
  with_groups(id, sample_n, 50)

d %>% 
  filter(id %in% sample_id) %>% 
  ggplot( aes( x=PA_lag, y=PA.std ) ) +
  geom_point(alpha=.5, shape=20) +
  geom_abline( aes(intercept=intercept, slope=slope, color=as.factor(id)),
               data = post2,
               alpha=.4) +
  geom_abline( aes(intercept=b_Intercept, slope=b_PA_lag),
               data = post[1:50, ],
               alpha=.4) +
  facet_wrap(~id) +
  guides(color="none") +
  labs(y="today",
       x="yesterday")
```

---

Back to the question: how much do these slopes differ?

```{r}
m6 %>% spread_draws(sd_id__PA_lag) %>% 
  mean_qi
```

---

## Correlations between parameters

Is someone's overall positive affect related to their holdover effect?


```{r}
m6 %>% spread_draws(cor_id__Intercept__PA_lag) %>% 
  mean_qi
```

---

```{r}
m6 %>% spread_draws(r_id[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_id) 
```

---

```{r}
m6 %>% spread_draws(r_id[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_id) %>% 
  ungroup() %>% 
  dplyr::select(.draw, Intercept, PA_lag) %>% 
  nest(data= -.draw) %>% 
  mutate(r = map_dbl(data, ~cor(.$Intercept, .$PA_lag))) 
```

---

```{r}
m6 %>% spread_draws(r_id[id, term]) %>% 
  pivot_wider(names_from = term, values_from = r_id) %>% 
  ungroup() %>% 
  dplyr::select(.draw, Intercept, PA_lag) %>% 
  nest(data= -.draw) %>% 
  mutate(r = map_dbl(data, ~cor(.$Intercept, .$PA_lag))) %>% 
  mean_qi(r)
```
