---
title: "Introduction to Bayes"
date: "2025-04-01"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sHcnk9KcoQVzDywxWRzVWHC"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "BIhz1dz5Vws", "Introduction", "1", TRUE,
  "f-sOSXyeeJE", "Data science and public service", "ds-ps", FALSE,
  "Z2foqSMWnGU", "Evidence, evaluation, and causation (1)", "evidence-evaluation-causation", FALSE,
  "W5ChuEhAzrs", "Evidence, evaluation, and causation (2)", "evidence-evaluation-causation1", FALSE,
  "D4LoEctZxHE", "Class details", "class-details", FALSE
)
```

## Session 1

## Before class

Nothing. But if you want to get familiar with the site or ee what I'm using as the basis for the first lecture, here you go:

- The [syllabus](/syllabus.qmd), [content](/content/index.qmd), [examples](/example/index.qmd), and [assignments](/assignment/index.qmd) pages for this class
- {{< fa book >}} [Introduction to Bayesian Inference for Psychology](/files/Etz and Vandekerckhove - 2018 - Introduction to Bayesian Inference for Psychology.pdf) [@etz_introduction_2018-1]
- {{< fa brands youtube >}} DJ Patil, ["What Makes a Radical and Revolutionary Technology?"](https://www.youtube.com/watch?v=UOoMzaWOQJA) 
    - <small>(DJ Patil is the former Chief Data Scientist of the United States under President Obama. He gave this forum address at Brigham Young University on February 13, 2018.)</small>

## During class

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/lectures/lecture01-1")
```

<iframe src="/lectures/lecture01-1.html" width="100%" height="500px"></iframe>

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::

