---
title: "Causal modeling"
date: "2025-04-15"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))
```


## Before class

- <i class="fab fa-youtube"></i> The Golum of Prague (51 min)


```{r, results = 'asis', echo = F}
library(vembedr)
embed_youtube("mBEA7PKDmiY")
```

## During class

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/lecture03-1")
```

<iframe src="/slides/lecture03-1.html" width="100%" height="500px"></iframe>

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
::: 