---
title: "Multivariate bananza"
date: "2025-05-22"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))
```


## Before class [OPTIONAL]

Turns out, `brms` can't do social networks. So we're going to skip this lesson in our course and replace it with something else -- specifically, multivariate outcomes and mixed effects location scale models. However, I know some of you may want to run social network analyses, so I'm keeping the video here for your reference. 

- <i class="fab fa-youtube"></i> Social networks (1 h 12 min)

```{r, results = 'asis', echo = F}
library(vembedr)
embed_youtube("hnYhJzYAQ60")
```


## During class

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/lecture08-1")
```
<iframe src="/slides/lecture08-1.html" width="100%" height="500px"></iframe> 
