---
title: Stan and rethinking
---

## Installing Stan and `rethinking`

You'll need to download a few packages in order to code starting in Week 2, and they're not as simple as going to CRAN. According to McElreath, there are 3 steps. (See here for his explaination, but I'll try to help here too):

### Install C++ toolchain

You might have one of these already. On a Mac, check by opening the Terminal app and running this code:

```
clang++ --version
```

If you get anything other than `command not found`, you're gold.

* On a Mac, download and install the app Xcode. 
* On a Windows, install C++ in [Visual Studio](https://learn.microsoft.com/en-us/cpp/build/vscpp-step-0-installation?view=msvc-170).

### Install `CmdStanR`

In R, run this code:

```{r, eval = F}
install.packages("cmdstanr", 
                 repos = c('https://stan-dev.r-universe.dev', getOption("repos")))

```

### Install `rethinking`

In R, run this code:

```{r, eval = F}
install.packages("shape")
install.packages("rethinking", 
                 repos=c(cran="https://cloud.r-project.org",
                         rethinking="http://xcelab.net/R"))
```

If you're asked whether you want to install from these sources (compilation of C/C++/Fortran), say yes.

### Install `brms` and `tidybayes`

While you're here, install these packages for later.

```{r, eval=F}
install.packages(c("brms", "tidybayes"))
```

